---
title: "Low microbial biomass within the reproductive tract of late postpartum dairy cows: a study approach - metadata and descriptive statistics"
author: "L. Lietaer"
output: html_notebook
---

This is a notebook about the metadata and descriptive statistics for the microbiome analysis data of Lietaer et al. (2020). 

When you execute code within the notebook, the results appear beneath the code. 


## Statistical analysis preface

Currently the following R packages were loaded

```{r, echo=FALSE}
#Import csv
if (!require("readr")) {
  install.packages("readr", dependencies = TRUE)
  library(readr)
}

#General purpose toolbox
if (!require("psych")) {
  install.packages("psych", dependencies = TRUE)
  library(psych)
}

#Dunn's Test - multiple comparisons using rank sums
if (!require("dunn.test")) {
  install.packages("dunn.test", dependencies = TRUE)
  library(dunn.test)
}

#Visualisations
if (!require("ggplot2")) {
  install.packages("ggplot2", dependencies = TRUE)
  library(ggplot2)
}

#Customizing 'ggplot2'- based publication ready plots
if (!require("ggpubr")) {
  install.packages("ggpubr", dependencies = TRUE)
  library(ggpubr)
}
```

## Data extraction, transformation and loading

The data was read from the csv file

```{r, echo=FALSE}
#import csv
data_descriptive_stats <- read_delim("./descriptive.csv", 
    ";", escape_double = FALSE, trim_ws = TRUE)

#summary statistics
summary(data_descriptive_stats)

#subset - no missing values qpcr
data_descriptive_stats2 <- data_descriptive_stats[-c(8,11,14,20,23,24),] 
#subset - controls excluded
data_descriptive_stats3 <- data_descriptive_stats[c(1:25),] 

#subset - vaginal samples excluded, no missing values qpcr
data_descriptive_stats4 <- data_descriptive_stats2[-c(17,18,19),] 
```

# Check normality - Histograms

```{r, echo=FALSE}
par(mfrow=c(3,2))
hist(data_descriptive_stats$Quantifluor)
hist(data_descriptive_stats$qPCR)
hist(data_descriptive_stats$Reads)
hist(data_descriptive_stats$Reads_decontam)
hist(data_descriptive_stats$OTUs)
hist(data_descriptive_stats$OTUs_decontam)
```
# Check normality - shapiro
```{r, echo=FALSE}
shapiro.test(data_descriptive_stats$Quantifluor)
shapiro.test(data_descriptive_stats$qPCR)
shapiro.test(data_descriptive_stats$Reads)
shapiro.test(data_descriptive_stats$Reads_decontam)
shapiro.test(data_descriptive_stats$OTUs)
shapiro.test(data_descriptive_stats$OTUs_decontam)
```
## Comparison tests - Wilcoxon test (two-level) or Dunn's Test (multi-level)
```{r, echo=FALSE}
#qPCR and Quantifluor: Uterus versus Controls

wilcox.test(qPCR ~ Ut_Control, data = data_descriptive_stats4,
                   alternative = "two.sided")

wilcox.test(Quantifluor ~ Ut_Control, data = data_descriptive_stats4,
                   alternative = "two.sided")
```

```{r, echo=FALSE}
#qPCR and Quantifluor: RT vs RB vs LT vs LB vs VAG

dunn.test(data_descriptive_stats3$qPCR, data_descriptive_stats3$Location, method ="bonferroni")

dunn.test(data_descriptive_stats3$Quantifluor, data_descriptive_stats3$Location, method ="bonferroni")
```

```{r, echo=FALSE}
#OTUs, Reads, qPCR and Quantifluor: sample (uterus - vagina) vs control

wilcox.test(OTUs ~ Sample_Control, data = data_descriptive_stats,
                   alternative = "two.sided")

wilcox.test(OTUs_decontam ~ Sample_Control, data = data_descriptive_stats,
                   alternative = "two.sided")

wilcox.test(Reads ~ Sample_Control, data = data_descriptive_stats,
                   alternative = "two.sided")

wilcox.test(Reads_decontam ~ Sample_Control, data = data_descriptive_stats,
                   alternative = "two.sided")

wilcox.test(qPCR ~ Sample_Control, data = data_descriptive_stats2,
                   alternative = "two.sided")

wilcox.test(Quantifluor ~ Sample_Control, data = data_descriptive_stats2,
                   alternative = "two.sided")
```

```{r, echo=FALSE}
#OTUs, Reads, qPCR and Quantifluor: uterus vs vagina

wilcox.test(OTUs ~ Ut_Vag, data = data_descriptive_stats3,
                   alternative = "two.sided")

wilcox.test(OTUs_decontam ~ Ut_Vag, data = data_descriptive_stats3,
                   alternative = "two.sided")

wilcox.test(Reads ~ Ut_Vag, data = data_descriptive_stats3,
                   alternative = "two.sided")

wilcox.test(Reads_decontam ~ Ut_Vag, data = data_descriptive_stats3,
                   alternative = "two.sided")

wilcox.test(qPCR ~ Ut_Vag, data = data_descriptive_stats3,
                   alternative = "two.sided")

wilcox.test(Quantifluor ~ Ut_Vag, data = data_descriptive_stats3,
                   alternative = "two.sided")
```

# Correlation between qPCR and quantifluor
```{r, echo=FALSE}
quanticor <- ggscatter(data_descriptive_stats2, x="qPCR", y="Quantifluor", 
          add = "reg.line", conf.int = TRUE, label = "ID",
          add.params = list(color = "dodgerblue4", fill = "lightgray"),
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "qPCR", ylab = "Quantifluor") +
  font("xlab", size = 15, color = "black", face = "bold")+
  font("ylab", size = 15, color = "black", face = "bold")
plot(quanticor)

cor.test(data_descriptive_stats2$qPCR, data_descriptive_stats2$Quantifluor, method="spearman")
```

```{r, echo=FALSE}
quanticor_samples <- ggscatter(data_descriptive_stats3, x="qPCR", y="Quantifluor", 
          add = "reg.line", conf.int = TRUE, label = "ID",
          add.params = list(color = "dodgerblue4", fill = "lightgray"),
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "qPCR", ylab = "Quantifluor") +
  font("xlab", size = 15, color = "black", face = "bold")+
  font("ylab", size = 15, color = "black", face = "bold")
plot(quanticor_samples)

cor.test(data_descriptive_stats3$qPCR, data_descriptive_stats3$Quantifluor, method="spearman")
```